---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%"
)
```

# predmicror

<!-- badges: start -->
<!-- badges: end -->

The `predmicror` is a collection of predictive microbiology models commonly fitted to experimental data. This first version includes a set of most popular full, reduced an no lag growth models.

## Installation

You can install the `predmicror` development version from [GitHub: https://github.com/fsqanalytics/predmicror](https://github.com/fsqanalytics/predmicror) using the `devtools` package:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("fsqanalytics/predmicror")
```

## Using predmicror

The easiest way to get started with `predmicror` is to work through the examples in the website and vignettes, where you can find information on:

* What your input data should look like
* How to fit a growth models
* How to fit a inactivation models
* How to fit a cardinal models

You can find the vignettes at the `predmicror` website [https://fsqanalytics.github.io/predmicror/](https://fsqanalytics.github.io/predmicror/). 

## Reporting bugs

You can **report bugs** at: [https://github.com/fsqanalytics/predmicror/issues](https://github.com/fsqanalytics/predmicror/issues).

