---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%"
)
```

# predmicror

[![R-CMD-check](https://github.com/fsqanalytics/predmicror/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fsqanalytics/predmicror/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/fsqanalytics/predmicror/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/fsqanalytics/predmicror/actions/workflows/pkgdown.yaml)

`predmicror` provides predictive microbiology model functions and convenience wrappers for fitting primary growth, microbial inactivation, and cardinal parameter models to experimental data.

The package can be used in two complementary ways:

- call the exported model equations directly inside `gslnls::gsl_nls()`;
- use the higher-level `fit_*()` wrappers for routine fitting, diagnostics, and model comparison.

## Installation

You can install the development version from GitHub:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("fsqanalytics/predmicror")
```

## Model families

`predmicror` includes models for three common predictive microbiology workflows.

| Workflow | Wrapper | Expected response scale | Examples |
|---|---|---|---|
| Primary growth | `fit_growth()` | natural logarithm, usually `lnN` | `HuangFM`, `BaranyiFM`, `RossoFM` |
| Microbial inactivation | `fit_inactivation()` | base-10 logarithm, usually `logN` | `WeibullM`, `WeibullPH`, `GeeraerdST` |
| Cardinal parameters | `fit_cardinal()` | square-root growth rate, usually `sqrtGR` | `CMTI`, `CMPH`, `CMAW`, `CMInh` |

List all models available through the fitting wrappers with:

```{r, eval = FALSE}
library(predmicror)

predmicror_models()
predmicror_models("growth")
predmicror_models("inactivation")
predmicror_models("cardinal")
```

## Quick start: fitting a primary growth model

```{r, eval = FALSE}
library(predmicror)

data(growthfull)

fit <- fit_growth(
  data = growthfull,
  model = "HuangFM",
  time = "Time",
  response = "lnN",
  start = list(Y0 = 0, Ymax = 22, MUmax = 1.7, lag = 5)
)

summary(fit)
coef(fit)
plot(fit)
```

## Diagnostics and model comparison

Fitted models returned by the wrappers are `predmicror_fit` objects. They support common S3 methods and helper functions for reporting.

```{r, eval = FALSE}
fit_metrics(fit)

aug <- predmicror_augment(fit)
head(aug)
```

Compare alternative candidate models with `compare_models()`:

```{r, eval = FALSE}
huang <- fit_growth(
  growthfull,
  model = "HuangFM",
  time = "Time",
  response = "lnN",
  start = list(Y0 = 0, Ymax = 22, MUmax = 1.7, lag = 5)
)

baranyi <- fit_growth(
  growthfull,
  model = "BaranyiFM",
  time = "Time",
  response = "lnN",
  start = list(Y0 = 0, Ymax = 22, MUmax = 1.7, lag = 5)
)

compare_models(HuangFM = huang, BaranyiFM = baranyi, sort_by = "AIC")
```

## Response scale conventions

The fitting wrappers do not transform the response automatically. Prepare the response variable before fitting:

- growth models: natural logarithm of microbial concentration, usually `lnN`;
- inactivation models: base-10 logarithm of microbial concentration, usually `logN`;
- cardinal models: square root of the growth rate, usually `sqrtGR`.

These conventions are important because parameter estimates depend on the response scale and on the time or environmental units used in the data.

## Documentation

More examples are available on the package website:

<https://fsqanalytics.github.io/predmicror/>

Useful articles include:

- model comparison workflows;
- microbial inactivation models;
- cardinal parameter models.

## Reporting bugs

Please report bugs and feature requests at:

<https://github.com/fsqanalytics/predmicror/issues>

## Citation

To cite `predmicror`, use:

```{r, eval = FALSE}
citation("predmicror")
```
