---
title: "Cardinal parameter models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cardinal parameter models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(predmicror)
```

Cardinal parameter models describe the effect of environmental factors on a
microbial growth response. In `predmicror`, the cardinal fitting wrapper expects
the response to be the square root of the growth rate, usually in a column named
`sqrtGR`.

This vignette shows how to fit cardinal models for temperature, pH, water
activity, and inhibitory compounds.

## Available cardinal models

```{r}
predmicror_models("cardinal")
```

## Robust fitting helper

Nonlinear fitting can depend on starting values. The helper below keeps the
vignette robust by reporting fitting problems without stopping the document
build.

```{r}
safe_fit <- function(expr) {
  withCallingHandlers(
    tryCatch(
      expr,
      error = function(e) {
        message("Model fit failed: ", conditionMessage(e))
        NULL
      }
    ),
    warning = function(w) {
      message("Model fit warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  )
}
```

## Temperature model

The `salmonella` dataset contains temperature and growth-rate data for
*Salmonella typhimurium* on cooked chicken.

```{r}
data(salmonella)
head(salmonella)
```

```{r}
temp_fit <- safe_fit(
  fit_cardinal(
    salmonella,
    model = "CMTI",
    x = "Temp",
    response = "sqrtGR",
    start = list(Tmax = 42, Tmin = 1, MUopt = 1, Topt = 37)
  )
)

temp_fit
```

```{r}
if (!is.null(temp_fit)) {
  fit_metrics(temp_fit)
}
```

```{r}
if (!is.null(temp_fit)) {
  grid <- data.frame(
    Temp = seq(min(salmonella$Temp), max(salmonella$Temp), length.out = 100)
  )
  pred <- predmicror_augment(temp_fit, newdata = grid)

  plot(
    sqrtGR ~ Temp,
    data = salmonella,
    xlab = "Temperature",
    ylab = "Square root growth rate",
    pch = 19
  )
  if (".fitted" %in% names(pred)) {
    lines(pred$Temp, pred[[".fitted"]], lwd = 2)
  }
}
```

## pH model

```{r}
data(ph)
head(ph)
```

```{r}
ph_fit <- safe_fit(
  fit_cardinal(
    ph,
    model = "CMPH",
    x = "pH",
    response = "sqrtGR",
    start = list(pHmax = 9.5, pHmin = 3.5, MUopt = 1, pHopt = 6.8)
  )
)

ph_fit
```

```{r}
if (!is.null(ph_fit)) {
  fit_metrics(ph_fit)
}
```

```{r}
if (!is.null(ph_fit)) {
  grid <- data.frame(pH = seq(min(ph$pH), max(ph$pH), length.out = 100))
  pred <- predmicror_augment(ph_fit, newdata = grid)

  plot(
    sqrtGR ~ pH,
    data = ph,
    xlab = "pH",
    ylab = "Square root growth rate",
    pch = 19
  )
  if (".fitted" %in% names(pred)) {
    lines(pred$pH, pred[[".fitted"]], lwd = 2)
  }
}
```

## Water activity model

```{r}
data(aw)
head(aw)
```

```{r}
aw_fit <- safe_fit(
  fit_cardinal(
    aw,
    model = "CMAW",
    x = "aw",
    response = "sqrtGR",
    start = list(AWmin = 0.90, MUopt = 1, AWopt = 0.995)
  )
)

aw_fit
```

```{r}
if (!is.null(aw_fit)) {
  fit_metrics(aw_fit)
}
```

```{r}
if (!is.null(aw_fit)) {
  grid <- data.frame(aw = seq(min(aw$aw), max(aw$aw), length.out = 100))
  pred <- predmicror_augment(aw_fit, newdata = grid)

  plot(
    sqrtGR ~ aw,
    data = aw,
    xlab = "Water activity",
    ylab = "Square root growth rate",
    pch = 19
  )
  if (".fitted" %in% names(pred)) {
    lines(pred$aw, pred[[".fitted"]], lwd = 2)
  }
}
```

## Inhibitor model

```{r}
data(inh)
head(inh)
```

```{r}
inh_fit <- safe_fit(
  fit_cardinal(
    inh,
    model = "CMInh",
    x = "Conce",
    response = "sqrtGR",
    start = list(MIC = max(inh$Conce) * 1.2, MUopt = 1, alpha = 1)
  )
)

inh_fit
```

```{r}
if (!is.null(inh_fit)) {
  fit_metrics(inh_fit)
}
```

```{r}
if (!is.null(inh_fit)) {
  grid <- data.frame(Conce = seq(min(inh$Conce), max(inh$Conce), length.out = 100))
  pred <- predmicror_augment(inh_fit, newdata = grid)

  plot(
    sqrtGR ~ Conce,
    data = inh,
    xlab = "Inhibitor concentration",
    ylab = "Square root growth rate",
    pch = 19
  )
  if (".fitted" %in% names(pred)) {
    lines(pred$Conce, pred[[".fitted"]], lwd = 2)
  }
}
```

## Collecting diagnostics

The fitted models above describe different environmental factors and should not
be ranked against each other using information criteria. Still, collecting their
metrics in one table is useful for reporting and quality control.

```{r}
cardinal_fits <- Filter(Negate(is.null), list(
  temperature = temp_fit,
  pH = ph_fit,
  water_activity = aw_fit,
  inhibitor = inh_fit
))

if (length(cardinal_fits) > 0L) {
  diagnostics <- do.call(
    rbind,
    Map(function(name, fit) {
      out <- fit_metrics(fit)
      out$workflow <- name
      out
    }, names(cardinal_fits), cardinal_fits)
  )
  rownames(diagnostics) <- NULL
  diagnostics
}
```

## Practical notes

Cardinal model parameters should be interpreted in the environmental units of
the input data. For example, `Tmin`, `Topt`, and `Tmax` are in the same units as
`Temp`, while `AWmin` and `AWopt` are in water-activity units. Always check that
the response scale is `sqrtGR` before fitting.
