---
title: "Fitting growth models using predmicror"
author:
   - "Vasco Cadavez"
   - "Ursula Gonzales-Barron"
date: 2022/04/30
output: 
  rmarkdown::html_vignette:
     number_sections: yes
     toc: yes
vignette: >
  %\VignetteIndexEntry{Fitting growth models using predmicror}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage{caption, amssymb, amsmath}
bibliography: predmicro.bib
csl: apa.csl
---

<!-- _Kellar and Lu, 2004_: [Modeling microbial responses in food](https://www.crcpress.com/Modeling-Microbial-Responses-in-Food/author/p/book/9780367394653) -->

In this tutorial we will explain how to fit primary growth models to experimental data. For this lab, we will use the `predmicror` R package developed to accommodate all the functions we are being using in predictive microbiology workshops. The `predmicror` package is outside the base R functions, and the first step is load the `predmicror`, package with the principal predictive microbial growth models, and `gslnls` which is a package to fit non-linear models using non-linear least squares.



```{r}
library(predmicror)
library(gslnls)
library(ggplot2)
```

## Loading data

To conduct statistical analyses, we need to import data into R working environment. The `predmicror` package has incorporated datasets, and we can use the `data()` function to load the example datasets to the working environment. Thus, we will start by fitting a full growth model to experimental data, thus we load the `growthfull.Rda` dataset which is part of the `predmicror` package.

```{r}
data(growthfull)
```

Now the dataset is available in the `R` environment, and we can print the entire `dataset` by typing `growthfull` or take a look to the first five lines by using the `head()`.

```{r}
growthfull
head(growthfull)
```

For data outer to `predmicror`package, usually in .csv format, which are flat text files we use the `read.csv()` function to import the CSV file into R environment. Before load a dataset, its good practice to assure that the dataset is located in the working directory, thus to import a CSV file called `growthfull.csv` into the R environment we do it with the next code section.

```{r, eval=FALSE}
growthfull <- read.csv("growthfull.csv", sep = ";", header = TRUE)
```

We have the dataset in the R environment, thus we can start checking the data proprieties. For example, the `str()` function gives information considering the structure of the variables (numeric, integer, etc.), and the `names()` function show us the variables names.

```{r}
str(growthfull)
names(growthfull)
```

## Plotting data

To check the relationship between `Time` and `lnN`, we might use the function `plot()`, and we can check the data by visual inspection.

```{r}
plot(lnN ~ Time,
  data = growthfull,
  xlab = "Time (hours)", ylab = "ln N",
  xlim = c(0, 20), ylim = c(0, 22)
)
```

To save the plot as an `.png` object, we can use the `png()` function.

```{r, eval=FALSE}
png("growthfull.png")
plot(lnN ~ Time,
  data = growthfull,
  xlab = "Time (hours)", ylab = "ln N",
  xlim = c(0, 20), ylim = c(0, 22)
)
dev.off()
```

## Fitting the Huang full model

We will start by fitting the _Huang_ full growth model to experimental data stored in the `growthfull` dataset. The Huang full growth model function (`Huang()`) from the `predmicror` package and this function will be fitted to data non-linear least squares function `gsl_nls()` implemented in the `gslnls` package.
First, we need a good guess for the starting values for the fitting procedure:

```{r}
start_values <- list(Y0 = 0.02, Ymax = 22, MUmax = 1.6, lag = 5)
```

Now we can fit the model to the data `growthfull`:

```{r}
fit <- gsl_nls(lnN ~ HuangFM(Time, Y0, Ymax, MUmax, lag),
  data = growthfull,
  start = start_values
)
```

Next, we can check the model parameters:

```{r}
summary(fit)
coef(fit)
```

The `predict()` function from the `gslnls` package can be used to produce both confidence and prediction intervals for the prediction of `lnN` for a given `Time`.


```{r}
newTimes <- data.frame(Time = seq(0, 24, by = 0.1))
fits <- data.frame(predict(fit, newdata = newTimes, interval = "confidence", level = 0.95))
str(fits)
preds <- data.frame(predict(fit, newdata = newTimes, interval = "prediction", level = 0.95))
str(preds)
```

Plot the observed data with the fitted values and confidence interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN",
  main = "Huang full model"
)
points(growthfull$Time, growthfull$lnN)
lines(newTimes$Time, fits$upr, col = "red")
lines(newTimes$Time, fits$lwr, col = "red")
```

Plot the observed data with the fitted values and prediction interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN", ylim = c(-1, 22), xlim = c(0, 24),
  main = "Huang full model"
)
points(growthfull$Time, growthfull$lnN)
lines(newTimes$Time, preds$upr, col = "red")
lines(newTimes$Time, preds$lwr, col = "red")
```


## Fitting the Fang no lag model

First, we want to load the data set:

```{r}
data(growthnolag)
growthnolag
```

Let's start with the Fang model, fit the model to the experimental data using nonlinear least squares function `gsl_nls()` implemented in the `gsl_nls` R package:

```{r}
start_values <- list(Y0 = 0.01, Ymax = 22, MUmax = 1.6)
fit <- gsl_nls(lnN ~ FangNLM(Time, Y0, Ymax, MUmax),
  data = growthnolag,
  start = start_values
)
fit
```

Next, we can check the model parameters:

```{r}
summary(fit)
```

Next, we can extract the model parameters and apply the model to new data

```{r}
newTimes <- data.frame(Time = seq(0, 18, by = 0.1))
fits <- data.frame(predict(fit, newdata = newTimes, interval = "confidence", level = 0.95))
str(fits)
preds <- data.frame(predict(fit, newdata = newTimes, interval = "prediction", level = 0.95))
```

Plot the observed data with the fitted values and confidence interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN",
  main = "Fang no lag model"
)
points(growthnolag$Time, growthnolag$lnN)
lines(newTimes$Time, fits$upr, col = "red")
lines(newTimes$Time, fits$lwr, col = "red")
```

Plot the observed data with the fitted values and prediction interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN", ylim = c(-1, 22), xlim = c(0, 18),
  main = "Fang no lag model"
)
points(growthnolag$Time, growthnolag$lnN)
lines(newTimes$Time, preds$upr, col = "red")
lines(newTimes$Time, preds$lwr, col = "red")
```

## Fitting the Buchanan reduced model

First, we want to load the data set:

```{r}
data(growthred)
growthred
```

Let's start with the Buchanan model, fit the model to the experimental data using nonlinear least squares function `gsl_nls()` implemented in the `gslnls` R package:

```{r}
start_values <- list(Y0 = 0.01, MUmax = 1.6, lag = 5)
fit <- gsl_nls(lnN ~ BuchananRM(Time, Y0, MUmax, lag),
  data = growthred,
  start = start_values
)
```

Next, we can check the model parameters:

```{r}
summary(fit)
```

Next, we can extract the model parameters and apply the model to new data

```{r}
newTimes <- data.frame(Time = seq(0, 16, by = 0.1))
fits <- data.frame(predict(fit, newdata = newTimes, interval = "confidence", level = 0.95))
str(fits)
preds <- data.frame(predict(fit, newdata = newTimes, interval = "prediction", level = 0.95))
```

Plot the observed data with the fitted values and confidence interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN",
  main = "Buchanan reduced model"
)
points(growthred$Time, growthred$lnN)
lines(newTimes$Time, fits$upr, col = "red")
lines(newTimes$Time, fits$lwr, col = "red")
```

Plot the observed data with the fitted values and prediction interval

```{r}
plot(newTimes$Time, fits$fit,
  type = "l", col = "blue",
  xlab = "time (days)", ylab = "logN", ylim = c(-1, 22), xlim = c(0, 16),
  main = "Buchanan reduced model"
)
points(growthred$Time, growthred$lnN)
lines(newTimes$Time, preds$upr, col = "red")
lines(newTimes$Time, preds$lwr, col = "red")
```
