---
title: "Microbial inactivation models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microbial inactivation models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(predmicror)
```

This vignette shows a complete workflow for fitting microbial inactivation
models with `predmicror`. Inactivation models in this package expect microbial
counts on the base-10 logarithmic scale, usually in a column named `logN`.

The workflow is:

1. inspect the available models;
2. fit one or more candidate models;
3. calculate model diagnostics;
4. extract fitted values and residuals;
5. generate predictions over a time grid.

## Available inactivation models

```{r}
predmicror_models("inactivation")
```

## Example data

The `mafart2005Li11` dataset contains inactivation data with time and log10
microbial counts.

```{r}
data(mafart2005Li11)
head(mafart2005Li11)
summary(mafart2005Li11)
```

```{r}
plot(
  logN ~ Time,
  data = mafart2005Li11,
  xlab = "Time",
  ylab = expression(log[10](N)),
  pch = 19
)
```

## Fitting candidate models

Nonlinear fitting can be sensitive to starting values. The helper below keeps
this vignette robust across platforms by returning `NULL` if a candidate model
cannot be fitted with the selected starting values.

```{r}
safe_fit <- function(expr) {
  withCallingHandlers(
    tryCatch(
      expr,
      error = function(e) {
        message("Model fit failed: ", conditionMessage(e))
        NULL
      }
    ),
    warning = function(w) {
      message("Model fit warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  )
}
```

A simple starting point is to fit the Weibull-Mafart model.

```{r}
weibull <- safe_fit(
  fit_inactivation(
    mafart2005Li11,
    model = "WeibullM",
    time = "Time",
    response = "logN",
    start = list(Y0 = max(mafart2005Li11$logN), sigma = 3, alpha = 1)
  )
)

weibull
```

Additional candidate models can be fitted and compared when they converge.

```{r}
weibull_ph <- safe_fit(
  fit_inactivation(
    mafart2005Li11,
    model = "WeibullPH",
    time = "Time",
    response = "logN",
    start = list(Y0 = max(mafart2005Li11$logN), k = 0.3, alpha = 1)
  )
)

huang_rgs <- safe_fit(
  fit_inactivation(
    mafart2005Li11,
    model = "HuangRGS",
    time = "Time",
    response = "logN",
    start = list(Y0 = max(mafart2005Li11$logN), k = 0.3, M = 1)
  )
)

fits <- Filter(Negate(is.null), list(
  WeibullM = weibull,
  WeibullPH = weibull_ph,
  HuangRGS = huang_rgs
))

names(fits)
```

## Diagnostics and model comparison

Use `fit_metrics()` for one fitted model and `compare_models()` for a ranked
summary of several fitted models.

```{r}
if (length(fits) > 0L) {
  fit_metrics(fits[[1]])
}
```

```{r}
if (length(fits) > 1L) {
  compare_models(fits, sort_by = "AIC")
}
```

## Fitted values and residuals

`predmicror_augment()` adds fitted values and residuals to the original data.
This is useful for diagnostic plots and for exporting results.

```{r}
if (length(fits) > 0L) {
  aug <- predmicror_augment(fits[[1]])
  head(aug)
}
```

```{r}
if (length(fits) > 0L) {
  aug <- predmicror_augment(fits[[1]])

  if (all(c(".fitted", ".resid") %in% names(aug))) {
    plot(
      aug[[".fitted"]], aug[[".resid"]],
      xlab = "Fitted values",
      ylab = "Residuals",
      pch = 19
    )
    abline(h = 0, lty = 2)
  }
}
```

## Prediction over a time grid

The same augment helper can be used with `newdata` to obtain predictions over a
regular time grid.

```{r}
if (length(fits) > 0L) {
  best_fit <- fits[[1]]
  if (length(fits) > 1L) {
    comparison <- compare_models(fits, sort_by = "AIC")
    best_fit <- fits[[comparison$model[1]]]
  }

  new_time <- data.frame(
    Time = seq(min(mafart2005Li11$Time), max(mafart2005Li11$Time), length.out = 100)
  )
  pred <- predmicror_augment(best_fit, newdata = new_time)

  plot(
    logN ~ Time,
    data = mafart2005Li11,
    xlab = "Time",
    ylab = expression(log[10](N)),
    pch = 19
  )
  if (".fitted" %in% names(pred)) {
    lines(pred$Time, pred[[".fitted"]], lwd = 2)
  }
}
```

## Practical notes

For inactivation models, verify the response scale before fitting. The wrappers
expect base-10 log counts. When fitting more complex models with residual tails,
use biologically plausible starting values and inspect parameter uncertainty;
residual population parameters can be weakly identified if the data do not show
a clear tail.
